---
title: Coursera - Reproducible Research - Assignment 2
subtitle: MORE SPECIFIC TITLE GOES HERE
author: Ron Ammar
output: 
  html_document:
    toc: true
    toc_depth: 4
---

```{r global_options, include=FALSE}
  # use include=FALSE to have the chunk evaluated, but neither the code nor its output displayed.
  knitr::opts_chunk$set(echo=TRUE, message=FALSE, fig.align="center",
                        fig.width=12, fig.height=8, fig.path='figure/',
                        dev='png')
```

```{r}
  # Clear the current session, to avoid errors from persisting data structures
  rm(list=ls())

  # Free up memory by forcing garbage collection
  invisible(gc())

  # Pretty printing in knitr
  library(printr)

  # Manually set the seed to an arbitrary number for consistency in reports
  set.seed(1234)

  # Do not convert character vectors to factors unless explicitly indicated
  options(stringsAsFactors=FALSE)

  startTime <- Sys.time()
```

## Background

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Purpose

The purpose of this analysis is to:

1. Complete the Reproducible Research course project #2.
2. Determine _____.

## Data Processing

**Analysis Date:** `r format(Sys.time(), "%k:%M:%S, %A, %B %d, %Y")` 

```{r, cache=TRUE}
  
  dat <- read.csv(bzfile("data/repdata-data-StormData.csv.bz2"))
```

### subsection 1

## Results

### subsection 1

------

## System Information

***Time required to process this report:*** *`r format(Sys.time() - startTime)`*

***R session information:***

```{r, echo_session_info}
  sessionInfo()
```
